---
title: "PS4"
author: "Joanna Zhang"
date: "3/02/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dendextend)
library(ggplot2)
library(dplyr)
```

## Performing k-Means By Hand
### 1. Plot the observations.
```{r}
x <- cbind(c(1, 1, 0, 5, 6, 4), c(4, 3, 4, 1, 2, 0))
plot(x, main = "Scatter Plot for Data Set X", xlab = "V1", ylab = "V2")
```

### 2. (5 points) Randomly assign a cluster label to each observation. Report the cluster labels for each observation and plot the results with a different color for each cluster (remember to set your seed first).
```{r}
set.seed(1405)
df <- as.data.frame(x) %>% 
  mutate(cluster = sample(2, nrow(x), replace =T))
df %>% 
  ggplot()+
  geom_point(aes(V1,V2, color = as.factor(cluster)))+
  theme_minimal()
```

### 3. (10 points) Compute the centroid for each cluster.
```{r}
centroids <- df %>% 
  group_by(cluster) %>% 
  summarize(mean_v1 = mean(V1), mean_v2 = mean(V2)) 
knitr::kable(centroid)
```


### 4. (10 points) Assign each observation to the centroid to which it is closest, in terms of Euclidean distance. Report the cluster labels for each observation.
```{r}
get_centroids <- function(data){
  curr_centroids <- data %>% 
    group_by(cluster) %>% 
    summarize(mean_v1 = mean(V1), mean_v2 = mean(V2))
  return(curr_centroids)
}

find_new_clstr <- function(data){
  curr_centroids <- get_centroids(data)
  newlabel <- vector()
  for (i in 1:nrow(data)){
  a <- data[[1]][i]
  b <- data[[1]][i]
  dist1 <- sqrt((a - curr_centroids[[2]][1])^2 + (b - curr_centroids[[3]][1])^2)
  dist2 <- sqrt((a - curr_centroids[[2]][2])^2 + (b - curr_centroids[[3]][2])^2)
  newlabel <- append(newlabel,ifelse(dist1 < dist2, 1, 2))
}
data <- data %>% 
  mutate(cluster = newlabel)  %>% 
  select(V1, V2, cluster)
return(data)
}
df<-find_new_clstr(df)
knitr::kable(df)

df %>% 
  ggplot()+
  geom_point(aes(V1,V2, color = as.factor(cluster)))+
  theme_minimal()
```


### 5. (5 points) Repeat (3) and (4) until the answers/clusters stop changing.
```{r}
converge = F
while(!converge){
  current_clstr <- df$cluster
  df<-find_new_clstr(df)
  converge = all(current_clstr == df$cluster)
}
```



### 6. (10 points) Reproduce the original plot from (1), but this time color the observations according to the clusters labels you obtained by iterating the cluster centroid calculation and assignments.
```{r}
df %>% 
  ggplot()+
  geom_point(aes(V1,V2, color = as.factor(cluster)))+
  theme_minimal()
```

## Clustering State Legislative Professionalism
### 1.
```{r}
load("Data and Codebook/legprof-components.v1.0.RData")
legprof<-x
```

### 2. (5 points) Munge the data:

a. select only the continuous features that should capture a state legislature’s level of “professionalism” (session length (total and regular), salary, and expenditures); 

b. restrict the data to only include the 2009/10 legislative session for consistency; 

c. omit all missing values; 

d. standardize the input features; e. and anything else you think necessary to get this subset of data into workable form (hint: consider storing the state names as a separate object to be used in plotting later)

```{r}

```

